---
title: "Evolutionary effects of HOI on patterns of species coexistence"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
require(deSolve) ## for integrating ordinary differential equations
require(tidyverse) ## for efficient data manipulation & plotting
require(cowplot) ## for arranging plots in a grid
require(statmod)
theme_set(theme_bw()) ## set basic plotting theme

## error function
erf <- function(x) 2*pnorm(x*sqrt(2))-1

```



```{r}
plot_density <- function(dat) {
  dat %>%
    ggplot +
    geom_line(aes(x=time, y=n, colour=as.factor(species))) +
    scale_y_continuous(name="population density", limits=c(0, NA)) +
    theme(legend.position="none") +ggtitle(dat$interaction) %>%
    return
}
plot_snapshot <- function(dat, moment=0, limits=c(-1,1), res=1001) {
  S <- dat %>% pull(species) %>% max ## number of species
  traitaxis <- seq(limits[1], limits[2], l=res) ## sampling the trait axis
  snap <- dat %>% filter(time==moment) %>% dplyr::select(-time) ## time = moment
  traits <- expand.grid(species=1:S, trait=traitaxis) %>% as_tibble ## trait table
  traits["density"] <- 0 ## add column for population densities
  for (i in 1:S) {
    v <- snap %>% filter(species==i) %>% dplyr::select(n, m, sigma)
    traits$density[(traits$species==i)] <- v$n * ## trait normally distributed,
      dnorm(traits$trait[(traits$species==i)], v$m, v$sigma) ## times density
  }
  
  traits$density[traits$density<max(traits$density)/1e3] <- NA
  landscape <- tibble(trait=traitaxis, r=0)  %>% ## phenotype-specific growth rates
    mutate(r=1) %>% 
    mutate(r=ifelse(r<=0, NA, r)) %>%
    mutate(r=r*max(traits$density, na.rm=TRUE))
  ggplot(traits) + ## generate plot
    geom_line(aes(x=trait, y=density, colour=factor(species)), na.rm=TRUE) +
    geom_ribbon(aes(x=trait, ymin=0, ymax=density, fill=factor(species)),
                alpha=0.15, colour=NA) +
    scale_x_continuous(name="trait value", limits=limits) +
    scale_y_continuous(name="density", limits=c(0, NA)) +
    theme(legend.position="none") %>%
    return
}


plot_trait <- function(dat) {
  dat %>%
    ggplot +
    geom_ribbon(aes(x=time, ymin=m-sigma, ymax=m+sigma,
                    fill=as.factor(species)), alpha=0.15) +
    geom_line(aes(x=time, y=m, colour=as.factor(species))) +
    ylab("trait value") +
    theme(legend.position="none") %>%
    return
}

plot_all <- function(dat, moment=0, limits=c(-1,1), res=1001) {
  cowplot::plot_grid(plot_density(dat), plot_trait(dat),
                     plot_snapshot(dat, moment, limits, res),align="hv", ncol = 1) %>%
    return
}

organize_results <- function(sol, pars) {
  S <- length(pars$sigma) ## number of species
  dat <- sol %>% as.data.frame %>% as_tibble ## convert to tibble
  names(dat)[1] <- "time" ## name the first column "time"
  names(dat)[2:(S+1)] <- paste0("n_", 1:S) ## name abundance columns (n_k)
  names(dat)[(S+2):(2*S+1)] <- paste0("m_", 1:S) ## name trait mean columns
  dat <- dat %>%
    gather("variable", "v", 2:ncol(dat)) %>% ## normalize the data
    separate(variable, c("type", "species"), sep="_") %>%
    spread(type, v) %>% ## separate columns for densities n and trait means m
    dplyr::select(time, species, n, m) %>% ## rearrange columns
    mutate(species=as.integer(species), sigma=pars$sigma[species], omega=pars$omega,
           h2=pars$h2[species],interaction= pars$interaction) ## add params
  return(as_tibble(dat))}


IntravarCV <- function(dat){
  mus <- dat$m
  d <- diff(mus)
  cvs<- (sd(d)/mean(d))
  
  return(cvs)
}

# gaussian quadrature function for approximating triple integrals
#takes into following values :
#1. index: which keeps a track of i, j, or k index
#2. m: vector of mean trait values
#3. params : parameter lists
#4. points: no. of nodes/points over which the gaussian quadrature is approximated.

gausquad<-function(index,m,params,interaction,points){
  S<-length(params$sigma)
  temp2<-dat2<-array(dim=c(points))
  
  # if the supposed interaction model is for more than 2 species i.e., HOI then ->>
  if(interaction=="HOI"){
    H<-t<-array(dim=c(points,points,points))
    temp1<-dat1<-array(dim=c(points,points))
    
    #nodes or points in the abscissa where the integral will be evaluated numerically
    z1<-gauss.quad.prob(points, dist="normal", mu=m[index$i],sigma =params$sigma[index$i])$nodes # z
    z2<-gauss.quad.prob(points,dist="normal", mu=m[index$j],sigma =params$sigma[index$j])$nodes # z'
    z3<-gauss.quad.prob(points, dist="normal", mu=m[index$k],sigma =params$sigma[index$k])$nodes # z''
    
    #weights of the gaussian distribution given by mean trait value mu_i and its variance \sigma_i
    w1<-gauss.quad.prob(points, dist="normal",mu=m[index$i],sigma =params$sigma[index$i])$weights #pi(z)
    w2<-gauss.quad.prob(points,dist="normal", mu=m[index$j],sigma =params$sigma[index$k])$weights #pj(z')
    w3<-gauss.quad.prob(points, dist="normal",mu=m[index$k],sigma =params$sigma[index$j])$weights #pk(z'')
    
    #nested loop for the integral goes over z,z' and z'' as:
    for (i in 1: points){
      for (j in 1:points){
        
        temp1[i,j]<-sum((0.5*(tanh((z1[i]-z2[j])/params$w)+1))*(0.5*(tanh((z2[j]-z3)/params$w)+1))*
                          w1[i]*w2[j]*w3+
                          (0.5*(tanh((z1[i]-z2[j])/params$w)+1))*(0.5*(tanh((z1[i]-z3)/params$w)+1))*
                          w1[i]*w2[j]*w3)
        
        
        dat1[i,j]<-sum((z1[i]-m[index$i])*(0.5*(tanh((z1[i]-z2[j])/params$w)+1))*(0.5*(tanh((z2[j]-z3)/params$w)+1))*
                         w1[i]*w2[j]*w3 +
                         (z1[i]-m[index$i])*(0.5*(tanh((z1[i]-z2[j])/params$w)+1))*(0.5*(tanh((z1[i]-z3)/params$w)+1))*
                         w1[i]*w2[j]*w3)
        
      }
      temp2[i]<-sum(temp1[i,])
      dat2[i]<-sum(dat1[i,])
    }
    return(list(Hjk= sum(temp2), bjk = sum(dat2)) )
    
  }
  
  #if the supposed interaction model is pairwise in nature, then this follows ->>
  else if (interaction =="pairwise"){
    hij<-bij<-array(dim=c(points,points))
    # print(index)
    #nodes oir points in the abscissa where the integral will be evaluated numerically
    z1<-gauss.quad.prob(points, dist = "normal", mu=m[index$r],sigma =params$sigma[index$r])$nodes #z'
    z2<-gauss.quad.prob(points, dist = "normal", mu=m[index$l],sigma =params$sigma[index$l])$nodes #z''
    
    #weights of the gaussian distribution given by mean trait value mu_i and its variance \sigma_i
    w1<-gauss.quad.prob(points, dist = "normal", mu=m[index$r],sigma =params$sigma[index$r])$weights #pi(z')
    w2<-gauss.quad.prob(points, dist = "normal", mu=m[index$l],sigma =params$sigma[index$l])$weights #pj(z'')
    
    
    #for the pairwise model however there are only two species interacting and hence i and j
    #or in other words the integral goes over z and z'
    for (i in 1: points){
      temp2[i]<- sum( (0.5*( tanh((z1[i]-z2)/params$w)+1))*w1[i]*w2)
      dat2[i]<- sum((z1[i]-m[index$r])*(0.5*(tanh((z1[i]-z2)/params$w)+1))*w1[i]*w2)
    }
    return(list(jk= sum(temp2), bk = sum(dat2)) )
    #return(list(Hjk= 0, bjk =0 ) )
  }
  
}


#main function for population and evolutionary trait dynamics
#time: time intervals
# x: state values
#parms: parameter values
eqs<- function(time, x, parms){
  S<-parms$S
  n <- x[1:S] ## species densities
  m<-x[(S+1):(2*S)] 
  if(parms$interaction == "HOI"){
    hijk<-bijk<-array(dim=c(S,S,S))
    h<-b<-array(dim=c(S,S))
    for(i in 1:S){
      for(j in 1:S){
        for(k in 1:S){
          indices<-list(i=i,j=j,k=k) #indices used to go over z, z' and z''
          
          # print(indices)
          hat<-gausquad(index=indices,m=m, params = parms,interaction=parms$interaction, points =5) # gaussian quad function approximating the integrals
          hijk[i,j,k]<-round(hat$Hjk,3)
          bijk[i,j,k]<-round(hat$bjk,3)
        }
        h[i,j]<- (hijk[i,j,]%*%n)
        b[i,j]<- (bijk[i,j,]%*%n)  
        
      }
    }
    n[n<parms$threshold]<-0  #cut-off for n if they go below a threshold value of 1e-9
    
    dndt<- n*(sum(parms$g*(parms$theta^2-
                             2*m*parms$theta+parms$sigma^2+
                             m^2)*n)*2*h%*%n)/((sum(n)*sum(n)*sum(n))) - n*parms$g*(parms$theta^2-2*m*parms$theta+parms$sigma^2+m^2)
    
    dmdt<-parms$h2*sum(parms$g*(parms$theta^2-
                                  2*m*parms$theta+parms$sigma^2+
                                  m^2)*n)*2*b%*%n/(sum(n)*sum(n)*sum(n))-
      parms$h2*2*parms$g*parms$sigma^2*(m-parms$theta)
    
    return(list(c(dndt,dmdt)))
  }
  
  # if the interaction model is pairwise in nature
  else if(parms$interaction=="pairwise"){
    t_m<-B<-array(dim = c(S,S))
    for(r in 1:S){
      for(l in 1:S){
        indices<-list(r=r,l=l,k=0)
        temp<-gausquad(index=indices,m = m,params=parms,interaction=parms$interaction, points=7)
        
        t_m[r,l]<-temp$jk
        B[r,l]<-temp$bk
      }
    }
    
    
    n[n<parms$threshold]<-0 
    dndt1<- n*(sum(parms$g*(parms$theta^2-2*m*parms$theta+parms$sigma^2+ m^2)*n)*(2*t_m%*%n/(sum(n)*sum(n)))) -n*parms$g*(parms$theta^2-2*m*parms$theta+parms$sigma^2+m^2)
    
    #dndt<-0
    dmdt1<-parms$h2*sum(parms$g*(parms$theta^2-
                                  2*m*parms$theta+parms$sigma^2+
                                  m^2)*n)*2*B%*%n/(sum(n)*sum(n))-parms$h2*2*parms$g*parms$sigma^2*(m-parms$theta)
    
    return(list(c(dndt1,dmdt1)))
  }
}


```


We can now run the model for $10^7$ time units. The code below generates a plot, where the top panel shows population densities in time, the second the trait means in time, and the third the final trait distributions of the species.

First we show for pairwise species competition model

```{r pairwise}
#set.seed(54321) ##set random seed (for reproducibility)
# number of species
S <- 15
x0 <- rep(runif(S))
x0 <- x0/sum(x0)
sum(x0)

m<-runif(S,-0.5,0.5) # initial random trait values
sigma<-runif(S,0.002,0.005) #trait vairance
w=0.8# parameter that captures the strength of competition
h2<-runif(S,0.15,0.15) #heritabilies of mean trait value
g<-0.5*10^-4 #strength in the mortality curve
theta<-0 #optimum mortality value
interaction = "pairwise" #pairwise or HOI


pars<-list( S=S,h2=h2,w=w,sigma=sigma,theta=theta, g=g,interaction=interaction, threshold = 1e-9) #
ic <- c(x0,m) #initial density values
tmax<-1e8
times <- seq(0, tmax, length=1000)

ode(func=eqs, y=ic, parms=pars, times=times) %>% 
  organize_results(pars) %>% plot_all(tmax)
```









We now show for the HOI evolutionary model for time of $10^6$ time points for $5$ species and modest strength in competition and low strength in the mortality curve captured by $g$ in the following script.


```{r HOI, fig.align="center", fig.height=6.0, fig.width=4.5, message=FALSE, warning=FALSE, out.width="90%"}
   ## set random seed (for reproducibility)
# number of species
S <- 10
x0 <- rep(runif(S))
x0 <- x0/sum(x0)


m<-runif(S,-0.5,0.5) # initial random trait values
sigma<-runif(S,0.002,0.005) #trait vairance
w=0.8 # parameter that captures the strength of competition
h2<-runif(S,0.15,0.15) #heritabilies of mean trait value
g<-0.5*10^-4 #strength in the mortality curve
theta<-0 #optimum mortality value
interaction = "HOI" #pairwise or HOI


pars<-list( S=S,h2=h2,w=w,sigma=sigma,theta=theta, g=g,interaction=interaction, threshold = 1e-9) #
ic <- c(x0,m) #initial density values
tmax<-1e8
times <- seq(0, tmax, length=1000)

d<-ode(func=eqs, y=ic, parms=pars, times=times) %>% 
  organize_results(pars) %>% 
  plot_all(tmax)
d
```


